\documentclass{beamer}
\usetheme{Montpellier}
\usecolortheme{dolphin}

\usepackage{graphicx} %For jpg figure inclusion
\usepackage{times} %For typeface
\usepackage{epsfig}
\usepackage{color} %For Comments
\usepackage[all]{xy}
\usepackage{float}
\usepackage{subfigure} 
\usepackage{hyperref}
\usepackage{url}
\usepackage{parskip}
\usepackage{multirow}

%% Elena's favorite green (thanks, Fernando!)
\definecolor{ForestGreen}{RGB}{34,139,34}
%% Joe's Color.
\definecolor{JoesGold}{RGB}{204,102,0}
%% Henry's Color.
\definecolor{Teal}{RGB}{2,132,130}
% Uncomment this if you want to show work-in-progress comments
\newcommand{\comment}[1]{{\bf \tt  {#1}}}
% Uncomment this if you don't want to show comments
%\newcommand{\comment}[1]{}
\newcommand{\emcomment}[1]{\textcolor{ForestGreen}{\comment{Elena: {#1}}}}
\newcommand{\joecomment}[1]{\textcolor{JoesGold}{\comment{Joe: {#1}}}}
\newcommand{\hfcomment}[1]{\textcolor{Teal}{\comment{Henry: {#1}}}}
\newcommand{\todo}[1]{\textcolor{blue}{\comment{To Do: {#1}}}}
\newcommand{\clocode}[1]{{\texttt {#1}}}
\newcommand{\fast}[1]{\textcolor{green}{#1}}
\newcommand{\slow}[1]{\textcolor{red}{#1}}

\begin{document}
\title{Exploration of parallelization efficiency in the Clojure programming language}
\date{\today}

\begin{frame}
\frametitle{Exploration of parallelization efficiency in the Clojure programming language}
{\centering
Midwest Instruction and Computing Symposium\par
April 25, 2014\par
Henry Fellows, Joe Einertson, and Elena Machkasova\par
}
\end{frame}
%frame

\begin{frame}[fragile]
\frametitle{Introduction}
	Our project is a comparison of parallelism methods in the Clojure programming language.
	\begin{itemize}
	\item Relatively new language.
	\item Designed for efficent parallel operations.
	\item Recently added new parallel library.
	\end{itemize}
	Motivations.
	\begin{itemize}
	\item Interest in using Clojure as an educational tool.
	\item Using concurrency in functional language.
	\item Developing concurrent algorithms.
	\end{itemize}
	
\end{frame}
%frame

\begin{frame}
\frametitle{Table of contents}
\tableofcontents %[currentsection] I don't know why this doesn't work.} 
\end{frame}
%frame
\section{Overview of Clojure}

\begin{frame}[fragile]
\frametitle{Intro to Clojure}
	\begin{itemize}
  	 \item Clojure is a dialect of Lisp.
	 \item Runs on the Java Virtual Machine (JVM).
  	 \item First introduced in 2007 by Rich Hickey.
  	 \item Immutable data structures.
	 \item Built-in support for parallelism.
	 \end{itemize}
\end{frame}
%frame

\begin{frame}[fragile]
\frametitle{Functional Languages and Lisps}
	Functional Languages
	\begin{itemize}
	 \item Clojure is a functional language.
  	 \item Treat computation as the evaluation of functions.
  	 \item Functional languages avoid direct memory manipulation.
       \end{itemize}
	Lisp of a family of programming languages 
		\begin{itemize}
		 \item Lisp-1 
		 \item Common Lisp
		 \item Scheme
		 \item Racket
		 \item Clojure
		\end{itemize}
\end{frame}
%frame
\begin{frame}[fragile]
\frametitle{Prefix Notation}
Can be generalized to \clocode{(function arg1 ... argN)}.
	\begin{verbatim}
	(+ 2 3)
	=> 5
	\end{verbatim}
Basic function syntax: \clocode{(defn name [args] expr)}
	\begin{verbatim}
	(defn add1 [num] (+ num 1))
	(add1 3)
	=> 4
	\end{verbatim}
\end{frame}
%frame

\begin{frame}[fragile]
\frametitle{Vectors}
A type of collection in Clojure.
Accessing items by index is $O(\log n)$.
	\begin{verbatim}
	(get [2 7 4 9 5] 3)	
	=> 9
	\end{verbatim}
\end{frame}
%frame
\begin{frame}[fragile]
\frametitle{High Order Functions}
%\hfcomment{mention code-as-data}
 Functions can take functions as arguments.
	\begin{verbatim}
	(map add1 [0 1 2 3 4])
	 => [1 2 3 4 5]
	\end{verbatim}
Another high order function, \clocode{reduce}.
	\begin{verbatim}
	(reduce + [1 2 3])
	 => 6
	\end{verbatim}
The combination of \clocode{reduce} and \clocode{map}.
	\begin{verbatim}
	(reduce + (map sqrt [1 4 25]))
	 => 8
	\end{verbatim}
\end{frame}
%frame
%\begin{frame}[fragile]
%\frametitle{Lazy Evaluation}
%	\begin{itemize}
%	  \item Delaying evaluation until the value is needed.
%  	  \item Infinite sequences - so long as it is not all called.
%       \end{itemize}	
%	\begin{verbatim}
%	 (take 10 (range))
%	 => (0 1 2 3 4 5 6 7 8 9)
%	\end{verbatim}
%\end{frame}
%frame
\section{Clojure Concurrency}

\begin{frame}
\frametitle{Concurrency}
	\begin{itemize}
	 \item Most processors are now being built with multiple cores.
	 \item Concurrency is the execution of multiple computations simultaneously.
	 \item Programming concurrent programs is \textit{considered hard}.
	 \item Deadlocking: two tasks are waiting for resources that the other task holds.
	 \item Immutable data structures make concurrency easier.
	\end{itemize}	
\end{frame}
%frame
\begin{frame}
\frametitle{Parallel Computation in Clojure}
	Clojure has several methods of parallelism.
	\begin{itemize}
	\item \clocode{pmap} is one of the early methods of parallelism in Clojure.
	\item Reducers is a new library introduced in 2012.
	\end{itemize}
\end{frame}
%frame
\begin{frame}[fragile]
\frametitle{Pmap}
	\begin{itemize}
	 \item A parallel version of \clocode{map}.
	 \item Has the same syntax as \clocode{map}.
	 \item On a sufficiently large collection, it will create additional threads.
	\end{itemize}	
	\begin{verbatim} 
	(pmap add1 [0 1 2 3 4])
	=> [1 2 3 4 5]
	\end{verbatim}
\end{frame}
%frame
\begin{frame}[fragile]
\frametitle{Reducers}
	\begin{itemize}
	 \item Released by Rich Hickey in May 2012.
	 \item Built on Java's fork/join framework.
	 \item Reducers provides parallel higher-order functions, with the same names as their serial counterparts.
	 \item \clocode{r/fold} is used in place of reduce.
	\end{itemize}
\end{frame}
%frame
\begin{frame}[fragile]
\frametitle{Implementation of Reducers}
	\begin{itemize}
	\item All collections come with a traversal mechanism.
	\item All reducers functions (\clocode{r/map, r/filter}) except \clocode{r/fold} provide a recipe.
	\item \clocode{r/fold} causes the evaluation of all recipes attached to a collection in parallel.
	\item Fork/Join framework creates one thread per core (as reported by OS).
	\end{itemize}
	\begin{verbatim}
	(r/fold + (r/map sqrt [1 4 25]))
	 => 8
	\end{verbatim}
\end{frame}
%frame
\begin{frame}[fragile]
\frametitle{Test Structure}
	\begin{itemize}
	\item Computationally expensive operations on large sets of integers
	
	\end{itemize}
	Three tests:
	\begin{itemize}
	\item Count-primes
	\begin{verbatim}
	(reduce + (map (one-if-prime-else-zero [...])))
	\end{verbatim}
	\item Sum-primes
	\begin{verbatim}
	(reduce + (map (zero-if-composite-else-n [...])))
	\end{verbatim}
	\item Sum-sqrt
	\begin{verbatim}
	(reduce + (map (sqrt [...])))
	\end{verbatim}
	\end{itemize}
\end{frame}
%frame 

\begin{frame}
\frametitle{Test Sub-Structure}
\begin{table}
\begin{center}
\begin{tabular}{|l|l|}
\hline 
Name & Description \\
\hline
map + reduce & serial map, serial reduce \\
pmap + reduce & parallel map, serial reduce \\
map + r/fold & serial map, parallel reduce \\
pmap + r/fold & parallel map, parallel reduce\\
r/map + r/fold & reducers parallel map, parallel reduce\\
r/fold & parallel reduce\\
\hline
\end{tabular}
\end{center}
\caption{Configurations for our tests}\label{table:tests}
\end{table}
The \clocode{r/fold} configuration does not have a mapping phase: the test code was rewritten to make it work with a single reduce.
\end{frame}
%frame
%\begin{frame}
%\frametitle{Fermat Primality Test}
%Two of the tests focus on prime testing;
%	\begin{itemize}
%	\item  Probabilistic algorithm called the \emph{Fermat primality test}.
%	\item Two parameters: a number to test, and the number of trials to run.
%	\item The more trials, the more accurate the result.
%	\item We run it with five trials, and numbers on the order of one billion.
%	\item \emph{Chosen because it is computationally expensive.}
%	\end{itemize}
%\end{frame}

\begin{frame}
\frametitle{Data Sets}
Count-primes
	\begin{itemize}
	\item Collection is 100,000 random integers between 0 and 1 billion.
	\item repeated 100 times, with new data each time. 
	\end{itemize}
Sum-primes
	\begin{itemize}
	\item Collection is 10,000 random integers  between 0 and 1 billion.
	\item repeated 1000 times, with new data each time. 
	\end{itemize}
Sum-sqrt
	\begin{itemize}
	\item Collection is 10,000 random integers  between 0 and 1 billion.
	\item repeated 1000 times, with new data each time. 
	\end{itemize}
\end{frame}
%frame

%\begin{frame}
%\frametitle{Compare-sum-sqrt}
%	\begin{itemize}
%	\item Sums the integer square root of each value in the given collection.
%	\item Like prime testing, integer square root is computationally difficult and a good 'real world' example.
%	\item Collection is 10,000 random integers uniformly distributed between 0 and 1 billion.
%	\item repeated 1000 times, with new data each time. 
%	\end{itemize}
%\end{frame}
%frame

\begin{frame}
\frametitle{Test Enviroments}
\begin{itemize}
 \item an Intel i7 CPU, with 4 cores.
 \item an Intel i5 CPU, with 2 cores.
 \item an AMD FX-8350 CPU, with 8 cores.
\end{itemize}
\end{frame}
%frame

\begin{frame}
\frametitle{Sum-Primes Results}
\begin{table}
\begin{center}
\hspace*{-0.2in}  %%%%%%% to move the table left
\begin{tabular}{|l|c|c|c|c|c|c|}
\hline
Run & \multirow{2}{1.5cm}{reduce, \\ map} 
&  \multirow{2}{1.5cm}{reduce, \\ pmap} 
&  \multirow{2}{1.3cm}{r/fold,  \\ pmap} 
&   \multirow{2}{1.3cm}{r/fold, \\  map} 
& r/fold& \multirow{2}{1.3cm}{r/fold, \\  r/map}\\
& & & & &  & %%%% to force vertical lines in the second row
\\
\hline
i7 & 208.0 & \fast{66.4} & \fast{61.7} & 207.0 & \fast{57.2} &  \fast{54.6} \\
i5 & 279.3 & 250.6 & 284.3 & 280.8 & \fast{132.0} & \fast{131.0} \\
AMD & 266.9 & 225.1 & 248.4 & 275.5 & \fast{59.2} & \fast{63.6} \\
\hline
\end{tabular}
\end{center}
\caption{Sum-Primes averages (ms).}\label{table:sum-primes}
\end{table}
\end{frame}
%frame

\begin{frame}
\frametitle{Count-Primes Results}
\begin{table}[h!]
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|}
\hline
Run & \multirow{2}{1.5cm}{reduce, \\ map} 
&  \multirow{2}{1.5cm}{reduce, \\ pmap} 
&  \multirow{2}{1.3cm}{r/fold,  \\ pmap} 
&   \multirow{2}{1.3cm}{r/fold, \\  map} 
& r/fold
\\
& & & & &   %%%% to force vertical lines in the second row
\\
\hline
i7  & 2084.6 & \fast{604.5} & \fast{597.1} & 2065.7 & \fast{535.8}\\
i5 & 2802.8 & 2567.7 & 2585.6 & 2774.0 & \fast{1269} \\
AMD & 2662.2 & 2411.3 & 2426.6 & 2647.9 & \fast{557.6}\\
\hline
\end{tabular}
\end{center}
\caption{Count-Primes averages (ms).}\label{table:count-primes}
\end{table}
\end{frame}
%frame

\begin{frame}
\frametitle{Sum-Sqrt Results}

\begin{table}[h!]
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|}
\hline
Run & \multirow{2}{1.5cm}{reduce, \\ map} 
&  \multirow{2}{1.5cm}{reduce, \\ pmap} 
&  \multirow{2}{1.3cm}{r/fold,  \\ pmap} 
&   \multirow{2}{1.3cm}{r/fold, \\  map} 
& r/fold
\\
& & & & &   %%%% to force vertical lines in the second row
\\
\hline
i7 & 115.4 & 128.7 & 109.7 & \fast{28.6} & \fast{30.5}\\
i5 & 120.1 & \slow{401.3} & \slow{414.0} & \fast{60.0} & \fast{58.0} \\
AMD & 115.9 & \slow{359.5} & \slow{367.6} & \fast{32.8} & \fast{32.4}\\
\hline
\end{tabular}
\end{center}
\caption{Sum-Sqrt averages (ms).}\label{table:sum-sqrt}
\end{table}
\end{frame}
%frame

\section{Conclusion}
\begin{frame}
\frametitle{Pmap and Thread Thrashing}
	Pmap is unreliable.
	\begin{itemize}
	\item Running times ranging from close to the best parallel runs, to worse than serial.
	\item Close to 2.5 times slower than serial methods.
	\end{itemize}
Pmap creates too many threads.
	\begin{itemize}
	\item This causes \emph{thread thrashing}.
	\item The number of treads leads to excessive context switching.
	\item Causing the process to choke on its own overhead.
	\end{itemize}
\end{frame}
%frame

\begin{frame}
\frametitle{Reducers}
	\begin{itemize}
	\item Reducers is \emph{fast}, running 15\% faster than pmap, when pmap was working well.
	\item \clocode{r/fold} + \clocode{r/map}, runs as fast as the one step \clocode{r/fold}.
	\item Relatively reliable.
	\end{itemize}
\end{frame}
%frame

\begin{frame}
\frametitle{Environments}
Intel i7 
	\begin{itemize}
	\item Resistant to thread thrashing.
	\item Caused by hyper-threading?
	\end{itemize}
Intel i5
	\begin{itemize}
	\item Slowest machine tested
	\item Not resistant to thread thrashing.
	\end{itemize}
AMD Fx-8350
	\begin{itemize}
	\item Slightly resistant to thread thrashing.
	\item Does not scale as well.
	\item Due to micro-architecture?
	\end{itemize}
	
\end{frame}
%frame

\begin{frame}
\frametitle{Conclusion}
	There's a lot to look into;
	\begin{itemize}
	\item Thread balancing in reducers.
	\item Optimal thread management.
	\item The effects of CPU architecture on thread thrashing.
	\end{itemize}
We still want to continue on our main interest, parallel algorithm development in functional languages.
\par
The authors thank Jon Anthony for helpful discussions and methodology suggestions. 
\end{frame}

\end{document} 